Using Scipion for Stream Image Processing at Cryo-EM Facilities

### 2018-09-25 Review/answer_referee.tex

\end{reviewer}

\begin{reviewer}
\section*{Reviewer 2}

\point{The authors claim that the strength of their software is in its ability to process data while being collected; a point that is not well-addresses by other software. At the same time, they acknowledge that UCSFImage4, Focus, and Relion are software packages similar to Scipion capable of processing data in a high-throughput manner while new data is being acquired at the microscope. A clear differentiation between Scipion and these software packages was lacking in this manuscript. What advantages Scipion provide over the above mentioned software packages?}

\reply Since there is already a published manuscript describing \scipion, in the present manuscript  we decided to focus on \scipion features that have been specifically developed for working at EM facilities (e.g. streaming). The main reasons why \scipion is better suited than other packages for EM work is rooted in \scipion main design rather than the new facilities related features. Nevertheless, the reviewer is right in pointing out than a ``A clear differentiation between Scipion and these software packages was lacking in this manuscript''. Therefore, we have decided to add the following subsection to the manuscript

@@

\begin{quote}

\textbf{\scipion Design and Architecture}

We would like to finish this section commenting on some of \scipion architecture design decisions that differentiate \scipion from another image processing frameworks. At the software architectural level,  the greatest difference 
between \scipion and the other EM image processing packages mentioned in the introduction 
is that  \scipion pays special attention to define abstraction layers to simplify maintenance
and extensibility, and that follows a strict workflow approach. A few examples will illustrate the main advantages of this design.

To create a new protocol in Scipion, only a Python script needs to be developed. From this script, the system will discover the new protocol and will automatically build
a form and store the protocol related information in a database.
\scipion protocol developers must certainly know \scipion conventions
but they do not need to modify or even understand the existing code.
In  \scipion next version, the abstraction will be increased
since protocols will be implemented as plug-ins,
that is, independent small python scripts  that add functionality to the main
\scipion application. The plug-in approach allows parallel development. That is,  
since protocols are implemented independently from the main application
they can be developed and released in parallel by different teams. In this way, support for a new version of \textit{Relion} does not require a new \scipion release but
a release of the independent \textit{Relion} plug-in

\scipion follows a pure workflow approach and therefore, implements pipelines and describes them as a set of interconnected steps (or protocols). In a workflow approach, users (including microscope staff) can reasonably be expected to modify the paths and incorporate new steps using a GUI. On the contrary, software packages that do not make this separation produce workflows that, in general, are more challenging to create or modify for non programmers.

Another advantage of a clear separation between steps and relationships 
is the improvement of load valance, reproducibility and  provenance (process of tracing and recording the origins of data and its movement between protocols). The first step in a \scipion workflow execution
is to store in a database each of the protocols to be executed as well as how 
outputs of one protocol are attached to the inputs of another protocol. Then, a workflow engine analyzes the stored data and using as constrain the number of CPUs assigned by the user to the workflow, optimizes the execution, that is, try to execute as many protocols in parallel as possible. Obviously, the data stored in the database  have enough information to execute each step and therefore guaranties reproducibility and provenance.

\end{quote}

### 2018-09-25 Review/2.overall.tex
to home

In the following we describe the most important modifications made to \scipion to facilitate its deployment and execution in centralized facilities.

@@

%\scipion is a multilayered application, the first layer handles the interaction with the user (\emph{presentation layer}), the second saves and retrieves the data used by the application  (\emph{data layer}) and finally, the third layer (\emph{business layer}) process the data following the user instructions.  
\scipion stream implementation make extensive use of threads. %A thread is the smallest unit of processing that can be scheduled by an operating system. 
Using threads, a programmer can effectively create two or more tasks that run at the same time and share the protocol resources (e.g. memory). 
%is based on periodic updates of the protocol input and output data during the image processing. 
%Each protocol execution in \scipion is an independent process that can create processing threads. 
Each protocol that works in streaming mode has one of its threads monitoring input and updating output  data while other threads take care of the data processing. This clear separation between data access and data processing marks an important difference between running a script or using \scipion, since it facilitates a form of computing in which several jobs are executed  concurrently, instead of sequentially. Moreover, \scipion workflow engine allows users to modify the workflow on the fly without modifying a single line of code.
%In practice, the latter follows  a more general and extensible approach that allows users to modify the workflow on the fly without modifying a single code line. %Moreover, it also allows to quickly develop new workflows .

@@

We would like to finish this section stressing that despite the modifications introduced to \scipion data and process model, the framework is backward compatible and keeps providing full data provenance. In other words, \scipion provides detailed information about the origin of the data  and the process to generate it.

>>>>

In the following we describe the most important modifications made to \scipion to facilitate its deployment and execution in centralized facilities. The section ends with a very brief description of \scipion design and architecture that highlights the most interesting features of this image processing framework.

@@

%\scipion is a multilayered application, the first layer handles the interaction with the user (\emph{presentation layer}), the second saves and retrieves the data used by the application  (\emph{data layer}) and finally, the third layer (\emph{business layer}) process the data following the user instructions.  
\scipion stream implementation make extensive use of threads. %A thread is the smallest unit of processing that can be scheduled by an operating system. 
Using threads, a programmer can effectively create two or more tasks that run at the same time and share the protocol resources (e.g. memory). 
%is based on periodic updates of the protocol input and output data during the image processing. 
%Each protocol execution in \scipion is an independent process that can create processing threads. 
Each protocol that works in streaming mode has one of its threads monitoring input and updating output  data while other threads take care of the data processing. This clear separation between data access and data processing marks an important difference between running a script or using \scipion, since it facilitates a form of computing in which several jobs are executed  concurrently, instead of sequentially. %Moreover, \scipion workflow engine allows users to modify the workflow on the fly without modifying a single line of code.
%In practice, the latter follows  a more general and extensible approach that allows users to modify the workflow on the fly without modifying a single code line. %Moreover, it also allows to quickly develop new workflows .

@@

%It is worth mentioning that despite the modifications introduced to \scipion data and process model, the framework is backward compatible and keeps providing full data provenance.% In other words, \scipion provides detailed information about the origin of the data  and the process to generate it.

\subsection{\scipion Design and Architecture}

We would like to finish this section commenting on some of \scipion architecture design decisions that differentiate \scipion from another image processing frameworks. At the software architectural level,  the greatest difference 
between \scipion and the other EM image processing packages mentioned in the introduction 
is that  \scipion pays special attention to define abstraction layers to simplify maintenance
and extensibility, and that follows a strict workflow approach. A few examples will illustrate the main advantages of this design.

To create a new protocol in Scipion, only a Python script needs to be developed. From this script, the system will discover the new protocol and will automatically build
a form and store the protocol related information in a database.
\scipion protocol developers must certainly know \scipion conventions
but they do not need to modify or even understand the existing code.
In  \scipion next version, the abstraction will be increased
since protocols will be implemented as plug-ins,
that is, independent small python scripts  that add functionality to the main
\scipion application. The plug-in approach allows parallel development. That is,  
since protocols are implemented independently from the main application
they can be developed and released in parallel by different teams. In this way, support for a new version of \textit{Relion} does not require a new \scipion release but
a release of the independent \textit{Relion} plug-in

\scipion follows a pure workflow approach and therefore, implements pipelines and describes them as a set of interconnected steps (or protocols). In a workflow approach, users (including microscope staff) can reasonably be expected to modify the paths and incorporate new steps using a GUI. On the contrary, software packages that do not make this separation produce workflows that, in general, are more challenging to create or modify for non programmers.

Another advantage of a clear separation between steps and relationships 
is the improvement of load valance, reproducibility and  provenance (process of tracing and recording the origins of data and its movement between protocols). The first step in a \scipion workflow execution
is to store in a database each of the protocols to be executed as well as how 
outputs of one protocol are attached to the inputs of another protocol. Then, a workflow engine analyzes the stored data and using as constrain the number of CPUs assigned by the user to the workflow, optimizes the execution, that is, try to execute as many protocols in parallel as possible. Obviously, the data stored in the database  have enough information to execute each step and therefore guaranties reproducibility and provenance.

### 2018-09-25 Review/3.workflow_cnb.tex
to home

\begin{figure}
  \centering
      \includegraphics[width=0.75\textwidth]{images/diagram.pdf}
  \caption{Graphical representation of the data flow through the CNB facility. Green, blue and red lines represent the microscope private network, the \cnb local area network and the public network respectively.}
  \label{fig:cnbpipeline}

@@

Data sets are recorded using \epu. The microscope control system has a 70 TB NAS (Network-attached storage) unit for saving movies as they are obtained (we will refer to this NAS as \mnas). This storage unit is shared through a Samba share and connected to the private network using a 10Gb connexion. \scipion is executed on a Linux server which has two 1Gb NICs (network interface controllers) that connect it to the private network and the LAN (local area network) respectively (we will refer to this machine as \scipionbox). \scipionbox has 32 CPUs at 2.40GHz, 62 Gb of RAM memory and 2 \textit{Quadro M4000} GPUs. Finally, a second NAS with four 1Gb NICs is linked to the private, LAN and WAN networks (we will refer to this second NAS as \onas) so it can retrieve data from any computer and serve it to external users.

The data flow is as follows: (a) the microscope system stores movies in the \mnas, (b) when executing movie-alignment protocols \scipionbox reads the movies from the \mnas but does not store them locally, (c) \onas has access to \mnas and \scipionbox hard disks (d) \onas has a collection of scripts than may be used to incrementally backup \scipion project and the movies to a local or remote disk, or serve this information through the Internet and, (e) in parallel, periodically \scipionbox transfers to  \hserver reports that may be checked by users or staff with an HTML browser (see \url{http://nolan.cnb.csic.es/scipionbox/lastHTMLReport/}).

>>>>

\begin{figure}
  \centering
      \includegraphics[width=0.75\textwidth]{images/diagram.pdf}
  \caption{Graphical representation of the data flow through the CNB facility. Green, blue and red lines represent the microscope private network, the \cnb local area network and the public (WAN) network respectively.}
  \label{fig:cnbpipeline}

@@

Data sets are recorded using \epu. The microscope control system has a 70 TB NAS (Network-attached storage) unit for saving movies as they are obtained (we will refer to this NAS as \mnas). This storage unit is shared through a Samba share and connected to the microscope private network using a 10Gb connexion. \scipion is executed on a Linux server which has two 1Gb NICs (network interface controllers) that connect it to the private network and the LAN (local area network) respectively (we will refer to this machine as \scipionbox). \scipionbox has 32 CPUs at 2.40GHz, 62 Gb of RAM memory and 2 \textit{Quadro M4000} GPUs. Finally, a second NAS with four 1Gb NICs is linked to the private, LAN and WAN networks (we will refer to this second NAS as \onas) so it can retrieve data from any computer and serve it to external users.

The data flow is as follows: (a) the microscope system stores movies in the \mnas, (b) when executing movie-alignment protocols \scipionbox reads the movies from the \mnas but does not store them locally, (c) \onas has access to \mnas and \scipionbox hard disks (d) \onas has a collection of scripts than may be used to incrementally backup \scipion projects and the associated movies to a local or remote disk, or serve this information through the Internet, and (e) in parallel, periodically \scipionbox transfers to  \hserver reports that may be checked by users or staff with an HTML browser (see a report example at \url{http://nolan.cnb.csic.es/scipionbox/lastHTMLReport/}).

### 2018-09-25 Review/3.workflow_diamond.tex
to home

At present, most of developments described in this subsection are available in a developer's branch  at \url{https://github.com/I2PC/scipion/tree/release-1.1-headless-devel}. Some of them (e.g. the monitor protocol) will be integrated in the main \scipion repository while the more Diamond specific software will be published in an independent repository. [** Ask Pablo about headless]

>>>>

At present, most of the developments described in this subsection are available in a developer's branch  at \url{https://github.com/I2PC/scipion/tree/release-1.1-headless-devel}. Some of them (e.g. the monitor protocol) will be integrated in the main \scipion repository while the more Diamond specific software will be published in an independent repository. \scipion headless mode is already integrated in the main \scipion branch.

### 2018-09-25 Review/4.future.tex
to home

\section{Future Developments and Usage}

@@

\subsection{Usage}
In the context of free and open source projects it is difficult to create usage statistics that are accurate. But certainly, if we want \scipion to succeed, we need to 
know how frequent the different protocols are used. This is the reason why we have recently developed our \textit{Scipion Usage Data Collector} that monitors -if activated by the user- the usage of the different protocols and send -via HTTP protocol- the information to the developer's team. This information is anonymous and cannot be used to identify the original computer in which \scipion has been executed.

Using the HTTP header it is possible to guess the country in which \scipion has been installed. With these information we have created the usage map and table available at URL \url{http://scipion.i2pc.es/report_protocols/scipionUsage/} and  \url{http://scipion.i2pc.es/report_protocols/protocolTable/}. The data corresponds on ``global'' \scipion usage rather than ``facility'' usage but there 
is a one to one relationship between the  countries with highest number of \scipion projects and the countries with \scipion based EM facilities. The only exception to the rule is Canada where two of the original \scipion developers are setting up a new facility (McGill University).

Note that, in a effort to report real use and skip test projects, the number of projects in the map refers to the number of non empty \scipion projects updated more than once. That is, 
\scipion projects that have been opened in at least two different days. Another constraint is that developers computers are filtered out.

>>>>

\section{Future Developments and Usage Statistics}

@@

\subsection{Usage Statistics}
In the context of free and open source projects it is difficult to create usage statistics that are accurate. But certainly, if we want \scipion to succeed, we need to 
know how frequent the different protocols are used. This is the reason why we have recently developed our \textit{Scipion Usage Data Collector} that monitors -if activated by the user- the usage of the different protocols and send -via HTTP protocol- the information to the developer's team. This information is anonymous and cannot be used to identify the original computer in which \scipion has been executed.

Using the HTTP header it is possible to guess the country in which \scipion has been installed. With these information we have created the usage map and table available at URLs \url{http://scipion.i2pc.es/report_protocols/scipionUsage/} and  \url{http://scipion.i2pc.es/report_protocols/protocolTable/}. The data shows ``global'' \scipion usage rather than ``facility'' usage but there 
is a one to one relationship between the  countries with highest number of \scipion projects and the countries with \scipion based EM facilities. The only exception to the rule is Canada where two of the original \scipion developers are setting up a new facility (McGill University).

Note that, in a effort to report real use and skip test projects, the number of projects in the map refers to the number of non empty \scipion projects updated more than once. That is, 
\scipion projects that have been opened in at least two different days. Another constraint is that developers' computers are filtered out.
